---
title: "Chain Binomial COVID Changepoints"
author: "Andre Ehrlich"
date: "25/3/2023"
output:
  pdf_document: default
  html_document: default
---
  
```{r setup, include=FALSE}
library(R2WinBUGS)
library(coda)
library(outbreaks)
library(BRugs)
library(incidence)
library(bayesplot)
library(mcmcplots)


library(zoo)
library(TTR)
library(data.table)
library(lubridate)
library(ggplot2)

knitr::opts_chunk$set(cache = TRUE, echo = TRUE, message = FALSE, warning = FALSE)
```


# Overview of Analysis
Each student will analyse the COVID-19 case data for the period 1 Sep 2020 – 20 June 2021 for 3 European countries.
## A) Chain binomial model 
- we saw in the practical, applied to the Shanghai influenza data. 
- Note that the COVID data are typically the number of new cases and not the total number of cases (active set).
## B) Peicewise constant with 1 change point 
- Fit a more realistic model where the infection rate (and therefore also Rt) is piece wise constant with 1 change point that you select.
## C) Extend the model to multiple changepoints.
## D) Estimate change-point 
- Fit the model with 1 changepoint by estimating the time point of change. 
- The Stagnant example could be useful for this task.
## E) Estimate multiple unknown changepoints (Optional) 
- Try a combination of © and (d) fitting a model with multiple unknown change points.


```{r}
# 
# # WHO COVID GLOBAL DATA
# who_data <- read.table("~/aueb/Semester 2A/Epidemic Models (Demiris)/WHO-COVID-19-global-data.csv", header = T, sep = ",")
# 
# who_data$Date_reported <- as.Date(who_data$Date_reported)
# 
# # Starting Population for selected countries
# # population <- data.frame("country_code"=c(), "pop"=c())
# # population <- rbind(population, c("CZ", 10708981))
# # population <- rbind(population, c("UK", 12))
# # population <- rbind(population, c("BG", 3))
# # colnames(population) <- c("country_code", "pop")
# # population
# 
# ccountry <- "CZ"
# # country_data <- who_data[who_data$Country_code == ccountry,]
# # country_data
# n_pop <- 10708981#population[population$country_code == ccountry]
# n_pop#
# 
# 
# who_data_subset <- who_data[
#   which(who_data$Country_code == ccountry & who_data$Date_reported >= start_date & who_data$Date_reported <= stop_date),
# ]
# who_data_subset
# 
# plot(who_data_subset$Date_reported, who_data_subset$New_cases, type="l")
# # This Data has some strange periodicity likely due to uneven case reporting
# # WMA ??
# 
# # Reported Cases each day between start and stop date.
# new_cases <- who_data_subset$New_cases
# n_obs <- length(new_cases)
# 
# # Deterministic Assumption: After 8 days of infection, subject is removed
# avg_infectious_period <- 8
# new_removals <- c(rep(0, avg_infectious_period), new_cases)
# new_removals <- new_removals[0:n_obs]
# 
# # Modify data into a form suitable for BUGS
# data <- list("n_obs", "n_pop", "new_cases", "new_removals")
```

# ECDC Data 
```{r ecdc_data, echo = TRUE}

# Time period 
start_date <- as.Date("2020-09-01", format = "%Y-%m-%d" ) 
stop_date <- as.Date("2021-06-20", format = "%Y-%m-%d")

# Deterministic Assumption: After 8 days of infection, subject is removed
avg_infectious_period <- 8

# smoothing 
x_day_moving_average <- 7

# Load Data 
ecdc <- read.table("~/aueb/Semester 2A/Epidemic Models (Demiris)/ecdc_covid_data.csv", header = T, sep = ",")
ecdc$dateRep <- as.Date(ecdc$dateRep, format="%d/%m/%Y")
ecdc <- setDT(ecdc)
```


```{r, country_data}

prepare_country_data <- function(ccountry, changepoints){
  
  # ccountry = 'LU'
  # changepoints = my_changepoints$LU
  
  # Select country
  ecdc_subset <- ecdc[geoId == ccountry]
  
  # Chronological order 
  ecdc_subset <- ecdc_subset[order(dateRep)]
  
  #### MA 
  # interpolate na's 
  ecdc_subset$cases_narm = na.approx(ecdc_subset$cases)
  # 7-day average, round to integer
  ecdc_subset$cases_ma <- round(WMA(ecdc_subset$cases_narm, n=7))
  
  # Time Period cutoff 
  ecdc_subset <- ecdc_subset[dateRep >= start_date & dateRep <= stop_date, ]
  
  
  # file.path(getwd(), "cz_cases.png")
  {
    plot_path <- file.path(getwd(), paste0(ccountry, "_cases"))
    # png(file=plot_path, width=1000, height=600)
    par(mfrow=c(1,1))
    
    ggplot() + 
      # Time Series 
      geom_line(data=ecdc_subset, mapping=aes(x=as.Date(dateRep), y=cases,  color = "Reported")) +
      geom_line(data=ecdc_subset, mapping=aes(as.Date(dateRep), cases_ma, color="7DMA")) +
      # geom_line(data=ecdc_subset, mapping=aes(y=cases,  color = "Reported")) + 
      # geom_line(data=ecdc_subset, mapping=aes(y=cases_ma, color="7DMA")) +

      scale_color_manual(name = "Cases", values = c("Reported" = "darkblue", "7DMA" = "red")) +
  
      # Changepoints 
      # geom_vline(aes(xintercept=as.numeric(cz_cp1), color="Changepoint1")) +
      # geom_vline(aes(xintercept=as.numeric(cz_cp2), color="Changepoint2")) +
      geom_vline(xintercept=as.numeric(changepoints[1]), color="blue", linetype=4) +
      geom_vline(xintercept=as.numeric(changepoints[2]), color="blue", linetype=4) +
      # 
          
      # Labels 
      ylab('Reported Cases') +
      xlab('Date') +
      ggtitle(paste(ccountry, " Reported Cases"))

    ggsave( paste0(ccountry, "_cases.png"),dpi = 300)

    
  }
  
  # Specify Model Parameters & Data
  n_pop <- ecdc[ecdc$geoId == ccountry, popData2020][1]
  new_cases <- ecdc_subset$cases_ma
  n_obs <- length(new_cases)
  new_removals <- c(rep(0, avg_infectious_period), new_cases)[0:n_obs]
  model_data <- list("n_obs"=n_obs, "n_pop" = n_pop, "new_cases"=new_cases, "new_removals"=new_removals)
  # return(list(model_data, ecdc_subset))
  
  return(model_data)
}


```



```{r, data_config}


country_codes <- c("CZ", "DK", "LU")

# Czechia
my_changepoints <- data.frame(
  "CZ" = c(
  as.Date("2020-12-01", format = "%Y-%m-%d" ), # 90
  as.Date("2021-02-01", format = "%Y-%m-%d" )  # 155
),
  "DK" = c(
  as.Date("2020-10-15", format = "%Y-%m-%d" ), 
  as.Date("2021-02-15", format = "%Y-%m-%d" ) 
),
  "LU" = c(
  as.Date("2020-09-20", format = "%Y-%m-%d" ), 
  as.Date("2021-01-15", format = "%Y-%m-%d" ) 
)
)


date_to_cp <- function(mydate) {
  as.numeric(mydate - as.Date("2020-09-01", "%Y-%m-%d"))
}

cz_cp1 <- date_to_cp(my_changepoints$CZ[1])
cz_cp2 <- date_to_cp(my_changepoints$CZ[2])

dk_cp1 <- date_to_cp(my_changepoints$DK[1])
dk_cp2 <- date_to_cp(my_changepoints$DK[2])

lu_cp1 <- date_to_cp(my_changepoints$LU[1])
lu_cp2 <- date_to_cp(my_changepoints$LU[2])

cz_data <- prepare_country_data("CZ", my_changepoints$CZ)
# cz_data


# Denmark 

dk_data <- prepare_country_data("DK", my_changepoints$DK)
# dk_data


lu_data <- prepare_country_data("LU", my_changepoints$LU)
# lu_data



```


The following chain binomial model belongs to the broader class of stochastic discrete-time SIR models.

$$\begin{array}{rcl}
new\_cases_{t} & \sim & Bin \left (S_{t-1}, 1-e^{-\beta \frac{I_{t-1}}{N}} \right )\\
S_{t} & = & S_{t-1} - new\_cases_{t} \\
I_{t} & = & I_{t-1} + new\_cases_{t} - removals_t
\end{array}$$

where $$\beta$$ is the probability that a susceptible individual has infectious contact with an infected individual and  becomes infected.

Note that, if we write $$q = e^{\frac{-\beta}{N}}$$ and assume that infectious period is fixed and constant, we have a Reed-Frost model. A susceptible at time t-1 can remain susceptible by avoiding being infected by all infectives I_{t-1}, and the probability of avoiding being infected by one infective is q.

```{r no_change, echo=TRUE}

# Fit Chain Binomial with no change points

rfcb_nocp <- function(my_data){
  
  CBmodel <- function(){
    S0 <- n_pop - 1
    I0 <- 1
    
    p[1] <- 1-exp(-(beta*I0/n_pop))
    new_cases[1] ~ dbin(p[1],S0)
    S[1] <- S0 - new_cases[1]
    I[1] <- I0 + new_cases[1] - new_removals[1]
    
    for (t in 2:n_obs){
      p[t] <- 1-exp(-(beta*I[t-1]/n_pop))
      new_cases[t] ~ dbin(p[t],S[t-1])
      S[t] <- S[t-1]- new_cases[t]
      I[t] <- I[t-1] + new_cases[t] - new_removals[t]
    }
    
    # prior   
    beta ~ dlnorm(0,5)        
  }
  
  filename<- file.path(getwd(), "CBmodel.bug")
  write.model(CBmodel, filename)
  
  n_chains=3
  n_burnin=500
  n_iter=25000
  n_thin=50
  set.seed(1234)
  
  # Specify parameters to monitor:
  params <- c("beta")
  
  # Generate initial values for the parameters:
  inits = function(){
    list(beta=runif(1,0.05,0.07))
  }
  
  time.start_mcmc <- Sys.time()
  
  # Run MCMC in openBUGS
  mcmc_fit <- openbugs(
      my_data, 
      inits, 
      model.file = filename, 
      parameters.to.save = params, 
      # program="OpenBUGS",
      n.chains = n_chains, 
      n.iter = n_iter, 
      n.burnin = n_burnin, 
      n.thin = n_thin
  )
  
  time.end_mcmc <- Sys.time()
  duration_mcmc <- time.end_mcmc - time.start_mcmc

  mcmc_fit
}

```


```{r one_changepoint,  echo=FALSE}

# 1 changepoint 
rfcb_1cp <- function(my_data, cp1){
  my_data$k <- cp1
  CBmodel2 <- function(){
    S0 <- n_pop - 1
    I0 <- 1
    # k <- cp1
  
    
    p[1] <- 1-exp(-(beta[1]*I0/n_pop))
    new_cases[1] ~ dbin(p[1], S0)
    S[1] <- S0 - new_cases[1]
    I[1] <- I0 + new_cases[1] - new_removals[1]
    
    for (t in 2:n_obs){
  
      # select beta vector depending on change point     
      param[t] <- 1 + step(k - t - 1) 
      p[t] <- 1 - exp(-(beta[param[t]]*I[t-1]/n_pop))
      new_cases[t] ~ dbin(p[t],S[t-1])
      S[t] <- S[t-1]- new_cases[t]
      I[t] <- I[t-1] + new_cases[t] - new_removals[t]
    }
    
    # prior   
    for (j in 1:2){
      beta[j] ~ dlnorm(0, 5)        
    }
  }
  
  paste(getwd())
  
  filename2 <- file.path(getwd(), "CBmodel2.bug")
  write.model(CBmodel2, filename2)
  
  n_chains=3
  n_burnin=500
  n_iter=25000
  n_thin=50
  set.seed(1234)
  
  # Specify parameters to monitor:
  params <- c("beta")
  
  # Generate initial values for the parameters:
  inits = function(){
    list(beta=c(runif(1,0.05,0.07), runif(1,0.15,0.17)))
  }
  
  time.start_mcmc <- Sys.time()
  #this function will call bugs through R, the user must set the correct file, where the bugs #executable is inside
  #If you choose debug == TRUE the Winbugs will remain open, even after the sampling is finished
  #There is a possible error Error in file(con, "wb") : cannot open the connection which can be #ignored
  
  mcmc_fit2 <- openbugs(
      my_data, 
      inits, 
      model.file = filename2, 
      parameters.to.save = params, 
      # program="OpenBUGS",
      n.chains = n_chains, 
      n.iter = n_iter, 
      n.burnin = n_burnin, 
      n.thin = n_thin
  )
  
  time.end_mcmc <- Sys.time()
  
  duration_mcmc <- time.end_mcmc - time.start_mcmc

  mcmc_fit2 
}

```


# BUGS step()
# y <- step(x)
# y = 0 if step(x) < 0
# y = 1 if x >= 0

```{r many_changepoint,  echo=FALSE}

rfcb_2cp <- function(my_data, cp1, cp2){
  
  my_data$k <- cp1
  my_data$k2 <- cp2
  
  CBmodel3 <- function(){
    S0 <- n_pop - 1
    I0 <- 1
  
    # k <- cp1
    # k2 <- cp2
    
    p[1] <- 1-exp(-(beta[1]*I0/n_pop))
    new_cases[1] ~ dbin(p[1], S0)
    S[1] <- S0 - new_cases[1]
    I[1] <- I0 + new_cases[1] - new_removals[1]
    
    for (t in 2:n_obs){
      
      param[t] <- 1 + step(t - k - 1) + step(t - k2 -1)
      p[t] <- 1 - exp(-(beta[param[t]]*I[t-1]/n_pop))
      new_cases[t] ~ dbin(p[t],S[t-1])
      S[t] <- S[t-1]- new_cases[t]
      I[t] <- I[t-1] + new_cases[t] - new_removals[t]
    }
    
    # prior   
    for (j in 1:3){
      beta[j] ~ dlnorm(0, 5)        
    }
  }
  
  paste(getwd())
  
  filename3 <- file.path(getwd(), "CBmodel3.bug")
  write.model(CBmodel3, filename3)
  
  n_chains=5
  n_burnin=2000
  n_iter=25000
  n_thin=50
  set.seed(1234)
  
  # Specify parameters to monitor:
  params <- c("beta")
  
  # Generate initial values for the parameters:
  inits = function(){
    list(beta=c(runif(1,0.05,0.07), runif(1,0.05,0.07), runif(1,0.05,0.07)))
  }
  
  time.start_mcmc <- Sys.time()
  #this function will call bugs through R, the user must set the correct file, where the bugs #executable is inside
  #If you choose debug == TRUE the Winbugs will remain open, even after the sampling is finished
  #There is a possible error Error in file(con, "wb") : cannot open the connection which can be #ignored
  
  mcmc_fit3 <- openbugs(
      my_data, 
      inits, 
      model.file = filename3, 
      parameters.to.save = params, 
      # program="OpenBUGS",
      n.chains = n_chains, 
      n.iter = n_iter, 
      n.burnin = n_burnin, 
      n.thin = n_thin
  )
  
  time.end_mcmc <- Sys.time()
  
  duration_mcmc <- time.end_mcmc - time.start_mcmc

  mcmc_fit3
}

```


```{r learn_many_changepoint,  echo=FALSE}

rfcb_2cp_learn <- function(my_data, cp1, cp2, cp_margin=30){
  
  # Express confidence in change point via the margin parameter
  my_data$cp1_lb <- cp1 - cp_margin
  my_data$cp1_ub <- cp1 + cp_margin

  my_data$cp2_lb <- cp2 - cp_margin
  my_data$cp2_ub <- cp2 + cp_margin
  
  
  CBmodel4 <- function(){
    S0 <- n_pop - 1
    I0 <- 1
  
    p[1] <- 1-exp(-(beta[1]*I0/n_pop))
    new_cases[1] ~ dbin(p[1], S0)
    S[1] <- S0 - new_cases[1]
    I[1] <- I0 + new_cases[1] - new_removals[1]
    
    for (t in 2:n_obs){
      
      param[t] <- 1 + step(t - k - 1) + step(t - k2 - 1)
      p[t] <- 1 - exp(-(beta[param[t]]*I[t-1]/n_pop))
      new_cases[t] ~ dbin(p[t],S[t-1])
      S[t] <- S[t-1]- new_cases[t]
      I[t] <- I[t-1] + new_cases[t] - new_removals[t]
    }
    
    # prior   
    
    # Ro parameteres
    for (j in 1:3){
      beta[j] ~ dlnorm(0, 5)        
    }
    
  
    k ~ dunif(cp1_lb,cp1_ub)
    k2 ~ dunif(cp2_lb,cp2_ub)
  
    # prior on number of change points
    # n_points ~ dunif(3,8)
  
    # changepoint priors
    # for (k in 1:n_points){
    #   change_points[k] ~ dunif((k-1)*90, (k-1)*90+20)
    # }
  
  }
  
  paste(getwd())
  
  filename4 <- file.path(getwd(), "CBmodel4.bug")
  paste(filename4)
  write.model(CBmodel4, filename4)
  
  n_chains=5
  n_burnin=5000
  n_iter=100000
  n_thin=50
  set.seed(1234)
  
  # Specify parameters to monitor:
  params <- c("beta", "k", "k2")
  
  # Generate initial values for the parameters:
  inits = function(){
    list(beta=c(runif(1,0.05,0.07), runif(1,0.05,0.07), runif(1,0.05,0.07)), k=runif(1,my_data$cp1_lb,my_data$cp1_ub), k2=runif(1,my_data$cp2_lb,my_data$cp2_ub))
  }
  
  time.start_mcmc <- Sys.time()
  #this function will call bugs through R, the user must set the correct file, where the bugs #executable is inside
  #If you choose debug == TRUE the Winbugs will remain open, even after the sampling is finished
  #There is a possible error Error in file(con, "wb") : cannot open the connection which can be #ignored
  
  mcmc_fit4 <- openbugs(
      my_data, 
      inits, 
      model.file = filename4, 
      parameters.to.save = params, 
      # program="OpenBUGS",
      n.chains = n_chains, 
      n.iter = n_iter, 
      n.burnin = n_burnin, 
      n.thin = n_thin
  )
  
  time.end_mcmc <- Sys.time()
  
  duration_mcmc <- time.end_mcmc - time.start_mcmc

 mcmc_fit4 
}

```

```{r, fit-models}

fit_all <- function(my_data, cp1, cp2, country){
  
  # Fit all the Models
  print(country)
  
  print("Model 1")
  mcmc_fit <- rfcb_nocp(my_data)
  print(mcmc_fit, digits = 3)
  diagnostic_plots(mcmc_fit, country, model_num="1")

  print("Model 2")
  mcmc_fit2 <- rfcb_1cp(my_data,cp1)
  print(mcmc_fit2, digits = 3)
    diagnostic_plots(mcmc_fit2, country, model_num="2")

  print("Model 3")
  mcmc_fit3 <- rfcb_2cp(my_data, cp1, cp2)
  print(mcmc_fit3, digits = 3)
    diagnostic_plots(mcmc_fit3, country, model_num="3")

  print("Model 4")
  mcmc_fit4 <- rfcb_2cp_learn(my_data, cp1, cp2, cp_margin=30)
  print(mcmc_fit4, digits = 3)
  diagnostic_plots(mcmc_fit4, country, model_num="4")

  return(c(mcmc_fit, mcmc_fit2, mcmc_fit3, mcmc_fit4))
}

```


```{r diagnostic-plots}

#### DIAGNOSTICS
# # Produce html file with trace, density, and autocorrelation plots. The files are displayed in the default internet browser
# mcmcplot(mcmc_results4)
# diagnostic_plots(mcmc_fit4, country="country", model_num="lala")
diagnostic_plots <- function(mcmc_obj, country="country", model_num="lala"){
  plot(mcmc_obj, display.parallel = TRUE) # gives a summary plot of parameters and credible intervals
  
  mcmc_obj_list <- as.mcmc.list(mcmc_obj)
  
  
  fname <- file.path(getwd(), paste0(country, "_model", model_num, "_ac"))
  png(fname)
  autocorr.plot(mcmc_obj_list)
  dev.off()

  fname <- file.path(getwd(), paste0(country, "_model", model_num, "_density"))
  png(fname)
  denplot(mcmc_obj_list, parms = c("deviance","beta", "k", "k2"))
  dev.off()

  fname <- file.path(getwd(), paste0(country, "_model", model_num, "_trace"))
  png(fname)
  traplot(mcmc_obj_list, parms = c("deviance","beta", "k", "k2"))
  dev.off()

}
```


```{r, run-all}

# RUN EVERYTHING
lu_models <- fit_all(lu_data, lu_cp1, lu_cp2, "LU")
cz_models <- fit_all(cz_data, cz_cp1, cz_cp2, "CZ")
dk_models <- fit_all(dk_data, dk_cp1, dk_cp2, "DK")

save.image(file='learn_many_changepoint.RData')
```